Model Finding for Recursive Functions in SMT

### 2016-01-26 ijcar2016/body.tex
some work on eval

\newcommand\isanun{Isa\xspace}
\newcommand\isa{Ip\xspace}
\newcommand\isam{Ip-Mut\xspace}
\newcommand\leon{Leon\xspace}
\newcommand\leonm{Leon-Mut\xspace}

@@

%\ct{The paper used "recursive datatype" in some places and "inductive datatype" in others. I have replaced both by "algebraic datatype" which is more accurate for \cvc.
%("recursive datatype" is a more general notion and "inductive datatype" has a stronger meaning in higher-order logics.)}  
%% I agree --JB.

@@

Nunchaku is a new higher-order model finder designed to be integrated with
several proof assistants. The first version, 0.1, was released in January 2016
with support for (co)datatypes, (co)recursive functions and (co)inductive
predicates. Support for higher-order functions is planned for the next
release. We have developed a preliminary Isabelle frontend and are planning
further frontends for Coq, the TLA\textsuperscript{+} Proof System, and
possibly other systems.

Nunchaku is the spiritual successor to Nitpick for
Isabelle/HOL~\cite{blanchette-nipkow-2010}, but is developed as a standalone
OCaml program, with its own input language. Whereas Nitpick generates a
succession of problems where cardinalities of finite types grow at each step,
Nunchaku translates its input to one first-order logic program that targets
the finite model finding fragment of CVC4, including interpreted
(co)datatypes \cite{reynolds-blanchette-2015-codata}.
Using CVC4 also allows Nunchaku to provide efficient arithmetic
reasoning and to detect unsatisfiability in addition to satisfiability.
% and makes
% equality reasoning straightforward.

@@

{\looseness=-1
In this section, we evaluate both the overall impact of the translation
introduced in Section~\ref{sec:encoding} and the performance of individual SMT
techniques.
%
We gathered 602 benchmarks from three sources, which we will refer to as
\isanun, \isa and \leon. 
The first source consists of 357 benchmarks coming from\ajr{where Isa-Nun benchmarks come from, what they represent, are they expected to be unsat/sat}.
The second source consists of the 79
benchmarks from the IsaPlanner
suite~\cite{DBLP:conf/itp/JohanssonDB10} that do not contain higher-order
functions. These benchmarks have been used recently as challenge problems for a
variety of inductive theorem provers. They heavily involve
recursive functions and are limited to a theory of algebraic datatypes
with a signature that contains uninterpreted function symbols over these datatypes. 
The third source consists of 166
benchmarks from the Leon repository,\footnote{%Available at
\url{https://github.com/epfl-lara/leon/}} which were constructed from
verification conditions about simple Scala programs. These benchmarks also
heavily involve recursively defined functions over algebraic datatypes, 
but cover a wide variety of additional theories, including bit vectors, arrays, and
both linear and nonlinear arithmetic. All benchmarks are in definitional form
with respect to a set of well-founded functions.
A majority of the benchmarks in the latter two sets are unsatisfiable.
}

For each of the 245 benchmarks in the \isa and \leon sets, 
we considered up to three randomly selected
mutated forms of its conjecture $\psi$. In particular, we considered unique
formulas that are obtained as a result of swapping a subterm of $\psi$ at
one position with another of the same type at another position.
Note that benchmarks created in this way have a high likelihood of having
small, easy-to-find countermodels. In total, we considered 213 mutated forms of
theorems from \isa and 427 mutated forms of theorems from \leon. We will
call these sets \isam and \leonm, respectively.
Thus, our benchmark set consists of 1242 benchmarks (640 mutants plus the
original 602). 
Each of the 1242 benchmarks were generated in SMT lib version 2.5 format~\cite{smtlib25}. %ajr : although these don't require 2.5 features
For \isanun, \ajr{describe generation of smt2 benchmarks without encoding}
The others were translated to SMT lib version 2 by the Leon tool.

We additionally considered these 1242 benchmarks both before and after the translation~$\conv$.
For example, \isa contains 79 original benchmarks $\varphi$ and 79 translated
benchmarks $\conv(\varphi)$.
For the \isanun, \ajr{describe generation of smt2 benchmarks with encoding}.
All others were translated by \cvc 
by outputting the result of running the preprocessor from Section~\ref{ssec:cvc4} on each benchmark.

For solvers, we considered the SMT solver \ziii~\cite{de-moura-bjoerner-2008},
which runs heuristic methods for quantifier instantiation~\cite{MouraBjoerner07}
as well as methods for finding models for quantified formulas~\cite{GeDeM-CAV-09}.
We also considered five configurations of \cvc~\cite{barrett-et-al-2011},
which we will refer to as \cvcd, \cvcf, \cvcfe, and \cvcfm.
The configuration \cvcd runs heuristic 
and conflict-based techniques for quantifier instantiation~\cite{ReynoldsTinelliMoura14},
but does not include techniques for finding models.
The other configurations run the finite model
finding procedure due to Reynolds et al.\ \cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}.
The configuration \cvcfe additionally incorporates heuristic quantifier instantiation as described in Section 2.3 of~\cite{reynolds-et-al-2013},
and \cvcfm incorporates the fairness scheme for monotonic types as described in Section~\ref{ssec:cvc4}.

@@

\isanun  & 0 & 14 & 0 & 0 & 0 & 155 & 0 & {\win 156} & 0 & 155
\\
\isa & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0
\\
\isam & 0 & 43 & 0 & 0 & 0 & {\win 161} & 0 & {\win 161} & 0 & {\win 161}
\\
\leon  & 8 & 30 & 6 & 6 & 6 & {\win 54} & 6 & 53 & 6 & {\win 54}
\\
\leonm & 6 & 61 & 0 & 0 & 2 & 180 & 2 & {\win 181} & 2 & {\win 181}
\\[\jot]
Total & 14 & 148 & 6 & 6 & 8 & 550 & 8 & {\win 551} & 8 & {\win 551}
\end{tabular}
\caption{\,Number of \emph{sat} responses on benchmarks without and with $\conv$ translation}
\label{fig:sat}
\end{figure}

\begin{figure}[t]
\small
%\normalsize
%\small
\centering
\begin{tabular}{l@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\,\,}}
  & \multicolumn{2}{c@{\kern1.5em}}{\phantom{0}\ziiib}     & \multicolumn{2}{c@{\kern1.5em}}{\phantom{0}\cvcd}     
  & \multicolumn{2}{c@{\kern1.5em}}{\phantom{0}\cvcf}      & \multicolumn{2}{c@{\kern1.5em}}{\cvcfe} 
  & \multicolumn{2}{c@{\,\,}}{\cvcfm}
\\%[-1pt]
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
\\
\midrule
\isanun & {\win 41} & 40 & 28 & 34 & 26 & 39 & 26 & 39 & 26 & 39
\\
\isa & 14 & {\win 15} & {\win 15} & {\win 15} & 1 & {\win 15} & {\win 15} & {\win 15} & 1 & {\win 15}
\\
\isam & 54 & 55 & {\win 56} & {\win 56} & 13 & 55 & {\win 56} & 55 & 13 & 55
\\
\leon & 48 & 52 & {\win 58} & 52 & 12 & 53 & {\win 58} & 50 & 12 & 53
\\
\leonm & 94 & 107 & 106 & 107 & 39 & {\win 108} & 106 & {\win 108} & 39 & {\win 108}
\\[\jot]
Total & 251 & 269 & 263 & 264 & 91 & {\win 270} & 261 & 267 & 91 & {\win 270}
\end{tabular}
\caption{\,Number of \emph{unsat} responses on benchmarks without and with $\conv$ translation}
\label{fig:unsat}
\end{figure}

The results are summarized in Figures \ref{fig:sat} and \ref{fig:unsat}.
The benchmarks and more detailed results are available online.%
\footnote{\url{http://lara.epfl.ch/~reynolds/IJCAR2016-recfun/}}
The figures are divided into benchmarks triggering \emph{unsat} and \emph{sat}
responses and further into benchmarks before and after the translation $\conv$.
The raw evaluation data reveals no cases in which a solver answered
\emph{unsat} on a benchmark $\varphi$ and \emph{sat} on its
corresponding benchmark $\conv( \varphi )$, or vice versa.
This is consistent with our expectations and Theorem~\ref{thm:equi}, 
since these benchmarks contain only well-founded function definitions.

Figure~\ref{fig:sat} shows that for untranslated benchmarks (the ``$\varphi$''
columns), the number of \emph{sat} responses is very low across all
configurations. This confirms the shortcomings of existing SMT techniques for
finding models for benchmarks containing recursively defined functions.
%
The translation $\conv$ (the ``$\conv(\varphi)$'' columns) has a major
impact. \cvcf finds 550 of the 1242 benchmarks to be satisfiable,
including 6~benchmarks in the nonmutated \leon benchmark set. 
The two optimizations for finite model finding in \cvc (configurations \cvcfe and \cvcfm) 
led to a net gain of 1 satisfiable benchmark each with respect to \cvcf.
The performance of \ziiib for countermodels also improved dramatically, as it
finds 134 more benchmarks to be satisfiable, including 5 that are not solved by \cvcf.
%
We conclude that the translation $\conv$ enables SMT
solvers to find countermodels for conjectures involving recursively defined
functions whose definitions are admissible.

Moreover,
the translation $\conv$ helps all configurations for \emph{unsat} responses as well.
\ziii solves a total of 269 with the translation, whereas it solves only 251 without it.
Surprisingly,
the configuration \cvcf, which is not tailored for handling unsatisfiable benchmarks,
solves 270 unsat benchmarks overall, which is more than both \cvcd and \ziii.
These results suggest that the translation do not degrade the performance 
of SMT solvers for unsatisfiable problems involving recursive functions, 
and instead often improve their performance.
\ajr{Importance of the translation helping unsat in practice for Isabelle/Nunchaku, if any}

>>>>

\newcommand\isanun{Nunchaku-Mut\xspace}
\newcommand\ip{IsaPlanner\xspace}
\newcommand\ipm{\ip-Mut\xspace}
\newcommand\leon{Leon\xspace}
\newcommand\leonm{\leon-Mut\xspace}

@@

@@

Nunchaku is a new higher-order model finder designed to be integrated with
several proof assistants. The first version, 0.1, was released in January 2016
with support for (co)algebraic datatypes, (co)recursive functions and
(co)inductive predicates. Support for higher-order functions is planned for
the next release. We have developed a preliminary Isabelle frontend and are
planning further frontends for Coq, the TLA\textsuperscript{+} Proof System,
and possibly other systems.

Nunchaku is the spiritual successor to Nitpick for
Isabelle/HOL~\cite{blanchette-nipkow-2010}, but is developed as a standalone
OCaml program, with its own input language. Whereas Nitpick generates a
succession of problems where cardinalities of finite types grow at each step,
Nunchaku translates its input to one first-order logic program that targets
the finite model finding fragment of CVC4, including
(co)algebraic datatypes \cite{reynolds-blanchette-2015-codata}.
Using CVC4 also allows Nunchaku to provide efficient arithmetic
reasoning and to detect unsatisfiability in addition to satisfiability.
% and makes
% equality reasoning straightforward.

@@

In this section, we evaluate both the overall impact of the translation
introduced in Section~\ref{sec:encoding} and the performance of individual SMT
techniques.
%
We gathered 602 benchmarks from three sources, which we will refer to as
\ip, \leon, and \isanun:
\begin{itemize}
\item
\ip{} consists of the 79 benchmarks from the IsaPlanner
suite~\cite{DBLP:conf/itp/JohanssonDB10} that do not contain higher-order
functions. These benchmarks have been used recently as challenge problems for a
variety of inductive theorem provers. They heavily involve
recursive functions and are limited to a theory of algebraic datatypes
with a signature that contains uninterpreted function symbols over these datatypes.

\smallskip

\item
\leon{} consists of 166
benchmarks from the Leon repository,\footnote{%Available at
\url{https://github.com/epfl-lara/leon/}} which were constructed from
verification conditions about simple Scala programs. These benchmarks also
heavily involve recursively defined functions over algebraic datatypes, 
but cover a wide variety of additional theories, including bit vectors, arrays, and
both linear and nonlinear arithmetic.

\smallskip

\item
\isanun{} consists of 357 benchmarks originating from Isabelle/HOL. They
involve (co)recursively defined functions over (co)algebraic datatypes and
uninterpreted functions but no other theories. They were obtained by mutation
of negated Isabelle theorems, as was done for evaluating
Nitpick~\cite{blanchette-nipkow-2010}. Benchmarks created by mutation have a
high likelihood of having small, easy-to-find models.
\end{itemize}

The \ip{} and \leon{} benchmarks are expressed in SMT-LIB 2.5 and are in
definitional form with respect to a set of well-founded functions.
The Leon tool was used to generate SMT-LIB files.
A majority of these benchmarks are unsatisfiable.
For each of the 245 benchmarks,
we considered up to three randomly selected
mutated forms of its goal $\psi$. In particular, we considered unique
formulas that are obtained as a result of exchanging a subterm of $\psi$ at
one position with another of the same type at another position.
In total, we considered 213 mutated forms of
theorems from \ip and 427 mutated forms of theorems from \leon. We will
call these sets \ipm and \leonm, respectively. Each of these benchmarks exists
in two versions:\ without and with the $\conv$~translation. Problems with
$\conv$ were produced by running \cvc's preprocessor % (Section~\ref{ssec:cvc4})
on each benchmark.

For \isanun, the Isabelle Nunchaku frontend was used to generate thousands of
Nunchaku problems from Isabelle/HOL theory files involving lists, trees, and
other functional data structures. Nunchaku was then used to generate SMT-LIB
files, again in two versions:\ without and with the $\conv$~translation.
Problems requiring higher-order logic were discarted, since Nunchaku does not
yet support them, leaving 357 problems.

% Our entire benchmark set consists of 1242 benchmarks (640 mutants plus the
% original 602).

%Each of the 1242 benchmarks were generated in SMT lib version 2.5 format~\cite{smtlib25}. %ajr : although these don't require 2.5 features

%We additionally considered these 1242 benchmarks both before and after the translation~$\conv$.
%For example, \ip contains 79 original benchmarks $\varphi$ and 79 translated
%benchmarks $\conv(\varphi)$.
%For the \isanun, \ajr{describe generation of smt2 benchmarks with encoding}.

Among SMT solvers, we considered \ziii~\cite{de-moura-bjoerner-2008} and
\cvc~\cite{barrett-et-al-2011}.
\ziii runs heuristic methods for quantifier instantiation~\cite{MouraBjoerner07}
as well as methods for finding models for quantified formulas~\cite{GeDeM-CAV-09}.
For \cvc, we considered four configurations, referred
to as \cvcd, \cvcf, \cvcfe, and \cvcfm.
The configuration \cvcd runs heuristic 
and conflict-based techniques for quantifier instantiation~\cite{ReynoldsTinelliMoura14},
but does not include techniques for finding models.
The other configurations run the finite model
finding procedure due to Reynolds et al.\ \cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}.
The configuration \cvcfe additionally incorporates heuristic quantifier instantiation as described in Section 2.3 of~\cite{reynolds-et-al-2013},
and \cvcfm incorporates the fairness scheme for monotonic types as described in Section~\ref{ssec:cvc4}.

@@

\ip & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0
\\
\ipm & 0 & 43 & 0 & 0 & 0 & {\win 161} & 0 & {\win 161} & 0 & {\win 161}
\\
\leon  & 8 & 30 & 6 & 6 & 6 & {\win 54} & 6 & 53 & 6 & {\win 54}
\\
\leonm & 6 & 61 & 0 & 0 & 2 & 180 & 2 & {\win 181} & 2 & {\win 181}
\\
\isanun  & 0 & 14 & 0 & 0 & 0 & 155 & 0 & {\win 156} & 0 & 155
\\[\jot]
Total & 14 & 148 & 6 & 6 & 8 & 550 & 8 & {\win 551} & 8 & {\win 551}
\end{tabular}
\caption{\,Number of \emph{sat} responses on benchmarks without and with $\conv$ translation}
\label{fig:sat}
\end{figure}

\begin{figure}[t]
\small
%\normalsize
%\small
\centering
\begin{tabular}{l@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\,\,}}
  & \multicolumn{2}{c@{\kern1.5em}}{\phantom{0}\ziiib}     & \multicolumn{2}{c@{\kern1.5em}}{\phantom{0}\cvcd}     
  & \multicolumn{2}{c@{\kern1.5em}}{\phantom{0}\cvcf}      & \multicolumn{2}{c@{\kern1.5em}}{\cvcfe} 
  & \multicolumn{2}{c@{\,\,}}{\cvcfm}
\\%[-1pt]
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
\\
\midrule
\ip & 14 & {\win 15} & {\win 15} & {\win 15} & 1 & {\win 15} & {\win 15} & {\win 15} & 1 & {\win 15}
\\
\ipm & 54 & 55 & {\win 56} & {\win 56} & 13 & 55 & {\win 56} & 55 & 13 & 55
\\
\leon & 48 & 52 & {\win 58} & 52 & 12 & 53 & {\win 58} & 50 & 12 & 53
\\
\leonm & 94 & 107 & 106 & 107 & 39 & {\win 108} & 106 & {\win 108} & 39 & {\win 108}
\\
\isanun & {\win 41} & 40 & 28 & 34 & 26 & 39 & 26 & 39 & 26 & 39
\\[\jot]
Total & 251 & 269 & 263 & 264 & 91 & {\win 270} & 261 & 267 & 91 & {\win 270}
\end{tabular}
\caption{\,Number of \emph{unsat} responses on benchmarks without and with $\conv$ translation}
\label{fig:unsat}
\end{figure}

The results are summarized in Figures \ref{fig:sat} and \ref{fig:unsat}.
Bold indicates the maximum of a row.
The benchmarks and more detailed results are available online.%
\footnote{\url{http://lara.epfl.ch/~reynolds/IJCAR2016-recfun/}}
The figures are divided into benchmarks triggering \emph{unsat} and \emph{sat}
responses and further into benchmarks before and after the translation $\conv$.
The raw evaluation data reveals no cases in which a solver answered
\emph{unsat} on a benchmark $\varphi$ and \emph{sat} on its
corresponding benchmark $\conv( \varphi )$, or vice versa.
This is consistent with our expectations and Theorem~\ref{thm:equi}, 
since these benchmarks contain only well-founded function definitions.

Figure~\ref{fig:sat} shows that for untranslated benchmarks (the ``$\varphi$''
columns), the number of \emph{sat} responses is very low across all
configurations. This confirms the shortcomings of existing SMT techniques for
finding models for benchmarks containing recursively defined functions.
%
The translation $\conv$ (the ``$\conv(\varphi)$'' columns) has a major
impact. \cvcf finds 550 of the 1242 benchmarks to be satisfiable,
including 6~benchmarks in the nonmutated \leon benchmark set. 
The two optimizations for finite model finding in \cvc (configurations \cvcfe and \cvcfm) 
led to a net gain of one~satisfiable benchmark each with respect to \cvcf.
The performance of \ziiib for countermodels also improved dramatically, as it
finds 134 more benchmarks to be satisfiable, including 5 that are not solved by \cvcf.
%
We conclude that the translation $\conv$ enables SMT
solvers to find countermodels for conjectures involving recursively defined
functions whose definitions are admissible.

Moreover,
the translation $\conv$ helps all configurations for \emph{unsat} responses as well.
\ziii solves a total of 269 with the translation, whereas it solves only 251 without it.
Surprisingly,
the configuration \cvcf, which is not tailored for handling unsatisfiable benchmarks,
solves 270 \emph{unsat} benchmarks overall, which is more than both \cvcd and \ziii.
These results suggest that the translation do not degrade the performance 
of SMT solvers for unsatisfiable problems involving recursive functions, 
and instead often improve their performance. It would be interesting to try this
out in Sledgehammer \cite{blanchette-et-al-2013-smt} and to try Nunchaku as a proof tool.
